canaccessR: An open data product for analyzing transportation accessibility to employment and grocery stores in Canada's largest metropolitan areas.

### 2025-03-10 paper/Manuscript-Data-Package.Rmd
generate descriptive statistics
create a separate R file to generate the descriptive statistics used in the paper

keywords: Accessibility; public transit; open data products (ODPs); travel time; employment; grocery stores.
classoption:
  - Royal
  - times
bibliography: bibfile.bib
bibliographystyle: sageh
output:
  rticles::sage_article:
    keep_tex: yes
editor_options: 
  markdown: 
    wrap: 72
---

@@

The objective of this paper is to describe the {canaccesR} open data package.
Its main contents are a set of public transit travel time matrices (TTM) estimates to employment and groceries stores from the 12 largest Canadian metropolitan areas in 2019 and 2023.
The results refer to all the Dissemination Areas (DA) (the smallest publicly available geography provided by Statistics Canada [@governmentofcanadaDictionaryCensusPopulation2021a]) within each metropolitan area, yielding origin-destination pairs containing information on travel time by public transit, population, and total employment within those spatial units.
These estimates were created by leveraging expertise in data science, computer programming, and transportation accessibility using the {r5r} package [@pereiraR5rRapidRealistic2021].
We used public transit schedule, transport network, population, and business location data from different sources to estimate these TTM's datasets.

@@

The locations included in the data package comprise the 12 largest (population-wise) Census metropolitan areas (CMA's) based on the 2021 Canadian Census [@governmentofcanada2021CensusPopulation2021] [^2]. 
These locations are Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo, London, and Halifax.
We used four main data sources to construct the {canaccesR} data package: General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data.

@@

## Travel time matrices

The main contents of the {canaccessR} package are the travel time matrices.
<!-- These matrices are data frames with origin-destination pairs, travel time by public transit, and population and employment by Census Dissemination Area.  -->
These matrices (data frames) comprise the estimated travel time by transit from each origin and destination Dissemination Area pairs.
The matrices also contain identifiers, such as the unique origin and destination DA codes, region identifier (*e.g.*, "cal" for the region Calgary) and name, the year in which the travel times refers to, and the date and time of departure for a trip that originates at the Dissemination Area of origin.
Besides these information, the matrices feature the total population in the origin DA and the total employment in the destination DA (both according to the 2016 Census).
There are two matrices for each metropolitan area, one containing the travel times to jobs and the other to groceries stores, making it 20 datasets (the information from Toronto, Hamilton, and Waterloo are aggregated at the Greater Golden Horseshoe travel time matrices).

**HOW TO MAKE THE INFORMATION FROM THE TTM INTO SPATIAL DATA?**
The information can be turned into spatial data using the

@@

## Other contents

Other sets of data are also available at the {canaccessR} package besides the travel time matrices.
These are the boundaries, socio-economic and demographic data (e.g., population, number of dwellings, number of individuals below the Low Income Measure, etc.) of the selected CMA's, disaggregated by DA.
In addition, the package also contains aggregated population statistics (for the selected CMA's) and transit revenue and ridership data aggregated by regional and national scale.

>>>>

keywords: Public transit accessibility; open data products (ODPs); R data package; travel time matrices.
classoption:
  - Royal
  - times
bibliography: bibfile.bib
bibliographystyle: sageh
output:
  rticles::sage_article:
    keep_tex: yes
editor_options: 
  markdown: 
    wrap: 72
---

@@

The objective of this paper is to describe the {canaccesR} open data package.
Its main contents are a set of public transit travel time matrices (TTM) estimates to employment and groceries stores from the 12 largest Canadian metropolitan areas in 2019 and 2023.
The results are stored at the Dissemination Areas (DA) [^1] level, yielding origin-destination pairs containing information on travel time by public transit, population, and total employment within each metropolitan area.
These estimates were created by leveraging expertise in data science, computer programming, and transportation accessibility using the {r5r} package [@pereiraR5rRapidRealistic2021].
We used public transit schedule, transport network, population, and business location data from different sources to estimate these TTM's datasets.

[^1]: Dissemination Areas are the smallest publicly available spatial unit provided by Statistics Canada [@governmentofcanadaDictionaryCensusPopulation2021a].

@@

The locations included in the data package comprise the 12 largest (population-wise) Census metropolitan areas (CMA's) based on the 2016 Canadian Census [@governmentofcanada2016CensusPopulation2016] [^2]. 
These locations are Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo, London, and Halifax.
We used four main data sources to construct the {canaccesR} data package: General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data.

@@

<!-- ## Travel time matrices -->

The package contains the following contents: travel time matrices, socio-economic and demographic data disaggregated at the DA level, the CMA areas' boundaries and backgrounds for plotting the data spatially, and aggregated statistics.

The main contents of the {canaccessR} package are the travel time matrices.
<!-- These matrices are data frames with origin-destination pairs, travel time by public transit, and population and employment by Census Dissemination Area.  -->
These matrices (datasets) comprise the estimated travel time by transit from and to each origin and destination Dissemination Area pairs.
The datasets also feature the total population in the origin DA and the total employment in the destination DA.
Besides these information, the matrices contain the unique origin and destination DA codes, region and name identifiers, the year in which the travel times refers to, and the date and time of departure for a trip that originates at the Dissemination Area of origin.
There are two matrices for each metropolitan area, one containing the travel times to jobs and the other to groceries stores [^4].
These information can be turned into spatial data using the socio-economic and demographic datasets.

[^4]: The information from Toronto, Hamilton, and Waterloo are aggregated at the Greater Golden Horseshoe travel time matrices, making it 20 datasets in total.

The socio-economic and demographic data contain other information disaggregated at the DA level for each CMA. 
These refer to total population by age groups, number of dwellings, number of individuals below the Low Income Measure, etc.
The aggregated statistics refer to population aggregates (for the selected CMA's) and transit revenue and ridership data aggregated by regional and national scale.

@@

```{r}

### 2025-03-10 paper/Manuscript-Data-Package.tex
generate descriptive statistics
create a separate R file to generate the descriptive statistics used in the paper

\keywords{Accessibility; public transit; open data products (ODPs);
travel time; employment; grocery stores.}

@@

The objective of this paper is to describe the \{canaccesR\} open data
package. Its main contents are a set of public transit travel time
matrices (TTM) estimates to employment and groceries stores from the 12
largest Canadian metropolitan areas in 2019 and 2023. The results refer
to all the Dissemination Areas (DA) (the smallest publicly available
geography provided by Statistics Canada
\citep{governmentofcanadaDictionaryCensusPopulation2021a}) within each
metropolitan area, yielding origin-destination pairs containing
information on travel time by public transit, population, and total
employment within those spatial units. These estimates were created by
leveraging expertise in data science, computer programming, and
transportation accessibility using the \{r5r\} package
\citep{pereiraR5rRapidRealistic2021}. We used public transit schedule,
transport network, population, and business location data from different
sources to estimate these TTM's datasets.

@@

The locations included in the data package comprise the 12 largest
(population-wise) Census metropolitan areas (CMA's) based on the 2021
Canadian Census \citep{governmentofcanada2021CensusPopulation2021}
\footnote{We included Oshawa, Ontario, as part of the Greater Toronto
  Area (GTA) because of the former's proximity to the latter. We also
  included Abbotsford-Mission, British Columbia, as part of the
  Vancouver metropolitan area because of the former's proximity to a
  transit station on the region's West Coast Express commuter rail line.}.
These locations are Toronto, Montreal, Vancouver, Ottawa-Gatineau,
Calgary, Edmonton, Quebec City, Winnipeg, Hamilton,
Kitchener-Cambridge-Waterloo, London, and Halifax. We used four main
data sources to construct the \{canaccesR\} data package: General
Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI's Enhanced
Points of Interest, and Statistics Canada Census data.

@@

\subsection{Travel time matrices}\label{travel-time-matrices}

The main contents of the \{canaccessR\} package are the travel time
matrices. These matrices (data frames) comprise the estimated travel
time by transit from each origin and destination Dissemination Area
pairs. The matrices also contain identifiers, such as the unique origin
and destination DA codes, region identifier (\emph{e.g.}, ``cal'' for
the region Calgary) and name, the year in which the travel times refers
to, and the date and time of departure for a trip that originates at the
Dissemination Area of origin. Besides these information, the matrices
feature the total population in the origin DA and the total employment
in the destination DA (both according to the 2016 Census). There are two
matrices for each metropolitan area, one containing the travel times to
jobs and the other to groceries stores, making it 20 datasets (the
information from Toronto, Hamilton, and Waterloo are aggregated at the
Greater Golden Horseshoe travel time matrices).

\textbf{HOW TO MAKE THE INFORMATION FROM THE TTM INTO SPATIAL DATA?} The
information can be turned into spatial data using the

@@

\subsection{Other contents}\label{other-contents}

Other sets of data are also available at the \{canaccessR\} package
besides the travel time matrices. These are the boundaries,
socio-economic and demographic data (e.g., population, number of
dwellings, number of individuals below the Low Income Measure, etc.) of
the selected CMA's, disaggregated by DA. In addition, the package also
contains aggregated population statistics (for the selected CMA's) and
transit revenue and ridership data aggregated by regional and national
scale.

>>>>

\keywords{Public transit accessibility; open data products (ODPs); R
data package; travel time matrices.}

@@

The objective of this paper is to describe the \{canaccesR\} open data
package. Its main contents are a set of public transit travel time
matrices (TTM) estimates to employment and groceries stores from the 12
largest Canadian metropolitan areas in 2019 and 2023. The results are
stored at the Dissemination Areas (DA) \footnote{Dissemination Areas are
  the smallest publicly available spatial unit provided by Statistics
  Canada \citep{governmentofcanadaDictionaryCensusPopulation2021a}.}
level, yielding origin-destination pairs containing information on
travel time by public transit, population, and total employment within
each metropolitan area. These estimates were created by leveraging
expertise in data science, computer programming, and transportation
accessibility using the \{r5r\} package
\citep{pereiraR5rRapidRealistic2021}. We used public transit schedule,
transport network, population, and business location data from different
sources to estimate these TTM's datasets.

@@

The locations included in the data package comprise the 12 largest
(population-wise) Census metropolitan areas (CMA's) based on the 2016
Canadian Census \citep{governmentofcanada2016CensusPopulation2016}
\footnote{We included Oshawa, Ontario, as part of the Greater Toronto
  Area (GTA) because of the former's proximity to the latter. We also
  included Abbotsford-Mission, British Columbia, as part of the
  Vancouver metropolitan area because of the former's proximity to a
  transit station on the region's West Coast Express commuter rail line.}.
These locations are Toronto, Montreal, Vancouver, Ottawa-Gatineau,
Calgary, Edmonton, Quebec City, Winnipeg, Hamilton,
Kitchener-Cambridge-Waterloo, London, and Halifax. We used four main
data sources to construct the \{canaccesR\} data package: General
Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI's Enhanced
Points of Interest, and Statistics Canada Census data.

@@

The package contains the following contents: travel time matrices,
socio-economic and demographic data disaggregated at the DA level, the
CMA areas' boundaries and backgrounds for plotting the data spatially,
and aggregated statistics.

The main contents of the \{canaccessR\} package are the travel time
matrices. These matrices (datasets) comprise the estimated travel time
by transit from and to each origin and destination Dissemination Area
pairs. The datasets also feature the total population in the origin DA
and the total employment in the destination DA. Besides these
information, the matrices contain the unique origin and destination DA
codes, region and name identifiers, the year in which the travel times
refers to, and the date and time of departure for a trip that originates
at the Dissemination Area of origin. There are two matrices for each
metropolitan area, one containing the travel times to jobs and the other
to groceries stores \footnote{The information from Toronto, Hamilton,
  and Waterloo are aggregated at the Greater Golden Horseshoe travel
  time matrices, making it 20 datasets in total.}. These information can
be turned into spatial data using the socio-economic and demographic
datasets.

The socio-economic and demographic data contain other information
disaggregated at the DA level for each CMA. These refer to total
population by age groups, number of dwellings, number of individuals
below the Low Income Measure, etc. The aggregated statistics refer to
population aggregates (for the selected CMA's) and transit revenue and
ridership data aggregated by regional and national scale.

@@
